\textbf{Unshuffling Permutations}

### 2016-04-07 slides/slides.tex
Typo
Suite

\usetheme{Boadilla}
\usefonttheme{serif}
\usecolortheme{default}

@@

\usepackage{euler,palatino}
\usepackage{amsmath,amssymb}
\usepackage[utf8]{inputenc}
\usepackage{xxcolor}
\usepackage{graphicx}
\usepackage[ruled]{algorithm2e}
\usepackage[english]{babel}
\usepackage{rotate}
\usepackage{shuffle}
%\usepackage{minted}
\usepackage{url}
\usepackage{etoolbox}
\usepackage{tikz}
\usepackage{rotating}
\usetikzlibrary{snakes}
\usetikzlibrary{trees}
\usetikzlibrary{arrows}
\usetikzlibrary{matrix}
\usetikzlibrary{calc}

@@

% Affichage du plan à chaque changement de partie.
\AtBeginSection[]{
    \frame<handout:0>{
        \frametitle{Plan}
        \tableofcontents[
            sectionstyle=show/hide,
            subsectionstyle=show/shaded/hide]}
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

@@

\uncover<3->{%
\begin{block}{Example}
    \vspace{-1em}
    \begin{equation*}\begin{split}
        \La\Lb \shuffle \textcolor{BrickRed}{\Lb\La} & =
        \La\Lb\textcolor{BrickRed}{\Lb\La}
        + \La\textcolor{BrickRed}{\Lb}\Lb\textcolor{BrickRed}{\La}
        + \La\textcolor{BrickRed}{\Lb\La}\Lb
        + \textcolor{BrickRed}{\Lb}\La\Lb\textcolor{BrickRed}{\La}
        + \textcolor{BrickRed}{\Lb}\La\textcolor{BrickRed}{\La}\Lb
        + \textcolor{BrickRed}{\Lb\La}\La\Lb \\
        &
        \uncover<4->{=
        2 \La\Lb\Lb\La + \La\Lb\La\Lb + \Lb\La\Lb\La + 2 \Lb\La\La\Lb}
    \end{split}\end{equation*}
    \vspace{-1.5em}
\end{block}}
\end{frame}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{???}
JE PENSE QU'IL FAUDRAIT PARLER ICI DES COUPLAGES PARFAITS POUR LES MOTS.
\medskip

ÇA PERMET DE FIXER QUELQUES DEFINITIONS POUR PREPARER LE TERRAIN POUR
NOTRE EXTENSION POUR LES PERMU DANS LA PARTIE ALGO.
\end{frame}

@@

There are several products on permutations, analogs of the shuffle
product of words. Among these, there are
\begin{itemize}
    \item the shifted shuffle product \BIB{Duchamp, Hivert, Thibon}{2002};
    \medskip

@@

\begin{block}{Main question}
    How the previous key results about the shuffle of words translate
    for the supershuffle of permutations?
\end{block}
\end{frame}

@@

\begin{block}{Definition}
    The \alert{unshuffling coproduct of permutations} is the
    coproduct on $\QQ[S]$ defined by
    \begin{equation*}
        \Delta(\pi) =
        \sum_{P \sqcup Q = \{1, \dots, |\pi|\}}
        \STD\left(u_{|P}\right) \otimes \STD\left(u_{|Q}\right).
    \end{equation*}
\end{block}

@@

\begin{block}{Example}
    The permutation
    \begin{math}
        \pi :=
        \textcolor{RoyalBlue}{25}\textcolor{BrickRed}{16}
        \textcolor{RoyalBlue}{7}\textcolor{BrickRed}{83}
        \textcolor{RoyalBlue}{4}
    \end{math}
    is square since $\pi$ can be obtained by shuffling
    $\textcolor{BrickRed}{1683}$ and $\textcolor{RoyalBlue}{2574}$.
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Some properties of square permutations}
The first numbers of square permutations are, size by size,
\begin{equation*}
    1, 0, 2, 0, 20, 0, 504, 0, 21032, 0, 1293418.
\end{equation*}
%\medskip

>>>>

\usetheme{Boadilla}
%\usetheme{default}
\usefonttheme{serif}
\usecolortheme{default}

@@

%\usepackage{euler,palatino}
\usepackage{lmodern}
\usepackage{amsmath,amssymb}
\usepackage[utf8]{inputenc}
\usepackage{xxcolor}
\usepackage{graphicx}
\usepackage[ruled]{algorithm2e}
\usepackage[english]{babel}
\usepackage{rotate}
\usepackage{shuffle}
%\usepackage{minted}
\usepackage{url}
\usepackage{etoolbox}
\usepackage{tikz}
\usepackage{rotating}
\usetikzlibrary{snakes}
\usetikzlibrary{trees}
\usetikzlibrary{arrows}
\usetikzlibrary{matrix}
\usetikzlibrary{calc}

@@

% Affichage du plan à chaque changement de partie.
\AtBeginSection[]{
    \frame<handout:0>{
        \frametitle{Contents}
        \tableofcontents[
            sectionstyle=show/hide,
            subsectionstyle=show/shaded/hide]}
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background}

@@

\uncover<3->{%
\begin{block}{Example}
    \vspace{-1em}
    \begin{equation*}\begin{split}
        \textcolor{RoyalBlue}{\La\Lb} \shuffle \textcolor{BrickRed}{\Lb\La}
        & =
        \textcolor{RoyalBlue}{\La\Lb}\textcolor{BrickRed}{\Lb\La}
        + \textcolor{RoyalBlue}{\La}\textcolor{BrickRed}{\Lb}%
        \textcolor{RoyalBlue}{\Lb}\textcolor{BrickRed}{\La}
        + \textcolor{RoyalBlue}{\La}\textcolor{BrickRed}{\Lb\La}%
        \textcolor{RoyalBlue}{\Lb}
        + \textcolor{BrickRed}{\Lb}\textcolor{RoyalBlue}{\La\Lb}%
        \textcolor{BrickRed}{\La}
        + \textcolor{BrickRed}{\Lb}\textcolor{RoyalBlue}{\La}%
        \textcolor{BrickRed}{\La}\textcolor{RoyalBlue}{\Lb}
        + \textcolor{BrickRed}{\Lb\La}\textcolor{RoyalBlue}{\La\Lb} \\
        &
        \uncover<4->{=
        2 \La\Lb\Lb\La + \La\Lb\La\Lb + \Lb\La\Lb\La + 2 \Lb\La\La\Lb}
    \end{split}\end{equation*}
    \vspace{-1.5em}
\end{block}}
\end{frame}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Square words}
\begin{block}{Definition}
    A word $u$ is a \alert{square} (w.r.t. the shuffle product
    $\shuffle$) if there is a word $v$ such that $u$ appears
    in~$v \shuffle v$.
\end{block}
\medskip

\begin{block}{Example}
    The word
    \begin{math}
        u := \textcolor{RoyalBlue}{3}\textcolor{BrickRed}{3}
        \textcolor{RoyalBlue}{12}\textcolor{BrickRed}{1}
        \textcolor{RoyalBlue}{2}\textcolor{BrickRed}{22}
    \end{math}
    is a square since $u$ can be obtained by shuffling $3122$ with itself.
\end{block}
\medskip

The first numbers of square binary words of length $2n$ are
\begin{equation*}
    1, 2, 6, 22, 82, 320, 1268, 5102, 20632, 83972.
\end{equation*}

\begin{block}{Open problem}
    Enumeration of square (binary) words.
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Recognizing square words}
\begin{block}{Definition}
    A \alert{perfect matching} of a word $u$ is a graph $(V, E)$ such
    that
    \begin{itemize}
        \item $V := \{(u_i, i) : i \in \{1, \dots, |u|\}$;
        \item $(u_i, i) \-- (u_j, j) \in E$ implies $u_i = u_j$;
        \item every vertex of $V$ belongs to exactly one edge of $E$.
    \end{itemize}
\end{block}

\begin{block}{Example}
    The word $21212233$ admits the perfect matching \vspace{-1em}
    \begin{equation*}
    \begin{split}
    \begin{tikzpicture}[xscale=.5,yscale=.3,inner sep=1pt,node distance=1.5cm]
        \tikzstyle{Vertex}=[draw,circle,minimum size=.35cm,font=\footnotesize]
        \draw node[Vertex,fill=RoyalBlue!30](U01){$(2, 1)$};
        \draw node[Vertex,fill=RoyalBlue!30,right of=U01](U02) {$(1, 2)$};
        \draw node[Vertex,fill=RoyalBlue!30,right of=U02](U03){$(2, 3)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U03](U04) {$(1, 4)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U04](U05) {$(2, 5)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U05](U06){$(2, 6)$};
        \draw node [Vertex,fill=RoyalBlue!30,right of=U06](U07){$(3, 7)$};
        \draw node [Vertex,fill=BrickRed!30,right of=U07](U08){$(3, 8)$};
        \draw [thick,>=latex']
        (U01.north) .. controls ($ (U01.north) + (0,2.5) $)
            and ($ (U06.north) + (0,2.5) $) .. (U06.north);
        \draw [thick,>=latex']
        (U02.north) .. controls ($ (U02.north) + (0,1.5) $)
            and ($ (U04.north) + (0,1.5) $) .. (U04.north);
        \draw [thick,>=latex']
        (U03.south) .. controls ($ (U03.south) + (0,-1.5) $)
            and ($ (U05.south) + (0,-1.5) $) .. (U05.south);
        \draw [thick,>=latex']
        (U07.north) .. controls ($ (U07.north) + (0,1.5) $)
            and ($ (U08.north) + (0,1.5) $) .. (U08.north);
    \end{tikzpicture}
    \end{split}\,.
    \end{equation*}
    \vspace{-1.5em}
\end{block}

\begin{block}{Definition}
    A perfect matching $(V, E)$ of a word $u$ is \alert{inclusion-free}
    if there are no edges $(u_i, i) \-- (u_j, j)$ and
    $(u_k, k) \-- (u_\ell, \ell)$ of $E$ such that $i < k < \ell < j$.
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Recognizing square words}

\begin{block}{Proposition \BIB{Rizzi, Vialette}{2013}}
    A word $u$ is a square iff $u$ admits an inclusion-free perfect
    matching.
\end{block}
%\medskip

\begin{block}{Example}
    The square word
    \begin{math}
        \textcolor{RoyalBlue}{3}\textcolor{BrickRed}{3}
        \textcolor{RoyalBlue}{12}\textcolor{BrickRed}{1}
        \textcolor{RoyalBlue}{2}\textcolor{BrickRed}{22}
    \end{math}
    admits the perfect matching \vspace{-1.5em}
    \begin{equation*}
        \begin{split}
        \begin{tikzpicture}[xscale=.5,yscale=.2,inner sep=1pt,node distance=1.5cm]
            \tikzstyle{Vertex}=[draw,circle,minimum size=.35cm,font=\footnotesize]
            \draw node[Vertex,fill=RoyalBlue!30](U01){$(3, 1)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U01](U02) {$(3, 2)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U02](U03){$(1, 3)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U03](U04) {$(2, 4)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U04](U05) {$(1, 5)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U05](U06){$(2, 6)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U06](U07){$(2, 7)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U07](U08){$(2, 8)$};
            \draw [thick,>=latex']
            (U01.north) .. controls ($ (U01.north) + (0,2.25) $)
                and ($ (U02.north) + (0,2.25) $) .. (U02.north);
            \draw [thick,>=latex']
            (U03.north) .. controls ($ (U03.north) + (0,2.25) $)
                and ($ (U05.north) + (0,2.25) $) .. (U05.north);
            \draw [thick,>=latex']
            (U04.north) .. controls ($ (U04.north) + (0,4.5) $)
                and ($ (U07.north) + (0,4.5) $) .. (U07.north);
            \draw [thick,>=latex']
            (U05.south) .. controls ($ (U05.south) + (0,-2.5) $)
                and ($ (U08.south) + (0,-2.5) $) .. (U08.south);
            \draw [thick,>=latex']
            (U06.north) .. controls ($ (U06.north) + (0,2.5) $)
                and ($ (U08.north) + (0,2.5) $) .. (U08.north);
        \end{tikzpicture}
        \end{split}
    \end{equation*}

    \vspace{-1.5em}
    which is inclusion-free.
\end{block}
%\medskip

\begin{block}{Example}
    The word $1221$ is not a square. Its only perfect matching\vspace{-1.5em}
    \begin{equation*}
        \begin{split}
        \begin{tikzpicture}[xscale=.5,yscale=.3,inner sep=1pt,node distance=1.5cm]
            \tikzstyle{Vertex}=[draw,circle,minimum size=.35cm,font=\footnotesize]
            \draw node[Vertex,fill=RoyalBlue!30](U01){$(1, 1)$};
            \draw node[Vertex,fill=RoyalBlue!30,right of=U01](U02) {$(2, 2)$};
            \draw node[Vertex,fill=BrickRed!30,right of=U02](U03){$(2, 3)$};
            \draw node [Vertex,fill=BrickRed!30,right of=U03](U04) {$(1, 4)$};
            \draw [thick,>=latex']
            (U01.north) .. controls ($ (U01.north) + (0,2.5) $)
                and ($ (U04.north) + (0,2.5) $) .. (U04.north);
            \draw [thick,>=latex']
            (U02.north) .. controls ($ (U02.north) + (0,1.5) $)
                and ($ (U03.north) + (0,1.5) $) .. (U03.north);
        \end{tikzpicture}\vspace{-1.5em}
        \end{split}
    \end{equation*}

    \vspace{-1.5em}
    is not inclusion-free.
\end{block}
\end{frame}

@@

There are several products on permutations, analogs of the shuffle
product of words. Among these, there are
\smallskip

\begin{itemize}
    \item the shifted shuffle product \BIB{Duchamp, Hivert, Thibon}{2002};
    \medskip

@@

\begin{block}{Main questions}
    \begin{itemize}
        \item How the previous key results about the shuffle of words
        translate for the supershuffle of permutations?
        \medskip

        \item
    \end{itemize}
\end{block}
\end{frame}

@@

\begin{block}{Definition}
    The \alert{unshuffling coproduct of permutations} is the
    coproduct on $\QQ[S]$ defined by
    \begin{equation*}
        \Delta(\pi) =
        \sum_{P \sqcup Q = \{1, \dots, |\pi|\}}
        \STD\left(\pi_{|P}\right) \otimes \STD\left(\pi_{|Q}\right).
    \end{equation*}
\end{block}

@@

\begin{block}{Example}
    The permutation
    \begin{math}
        \pi :=
        \textcolor{RoyalBlue}{25}\textcolor{BrickRed}{16}
        \textcolor{RoyalBlue}{7}\textcolor{BrickRed}{83}
        \textcolor{RoyalBlue}{4}
    \end{math}
    is square since $\pi$ can be obtained by shuffling
    $\textcolor{BrickRed}{1683}$ and $\textcolor{RoyalBlue}{2574}$
    and
    \begin{math}
        \STD(\textcolor{BrickRed}{1683})
        = \STD(\textcolor{RoyalBlue}{2574})
        = 1342
    \end{math}.
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame} \frametitle{Some properties of square permutations}
The first numbers of square permutations of size $2$ are
\begin{equation*}
    1, 2, 20, 504, 21032, 1293418.
\end{equation*}
%\medskip
